---
title: "Explore SF Salary Data"
author: "Michael Griffiths"
output:
  html_document:
    highlight: pygments
    toc: yes
  pdf_document:
    toc: yes
---

Kaggle Datasets provides data on[San Francisco city employee data](https://www.kaggle.com/kaggle/sf-salaries). 

For fun, let's do some EDA.

First, let's set our working directory
```{r}
setwd(dir = '../sf_salaries/')
```

```{r warning=F, comment=F, message=F}
library(dplyr)
library(RSQLite)
library(tidyr)
library(ggplot2)
library(readr)
library(stringr)
library(scales)
library(mixtools)
library(gender) # add package gender for gender prediction from first names
```

# Data

The files are in the `../output` subdirectory.

```{r ListData}
list.files("output/")
```

The two files are the SQLite database (`database.sqlite`), and the CSV file (`Salaries.csv`). In theory, these have exactly the same data.

Well, let's check.

```{r}
# The CSV
data_csv <- read_csv("output/Salaries.csv", na=c("Not Provided"))
(glimpse(data_csv))

# The database
db <- src_sqlite("output/database.sqlite")

# Get table list
# Preview tables
(glimpse(db %>% tbl("Salaries")))
```

A quick visual estimation suggest that they're very similar. A keen eye might see that in the CSV, field `OvertimePay` is inferred as `dbl`, but in the SQL database it's coerced to `int`. 

Of course - does that mean there are differences? If all the `OvertimePay` has zero cents, they should be.

```{r}
head(data_csv) == db %>% tbl("Salaries") %>% head
```

Well: nope, they're not the same. The good news is that's the only column that has any differences in the first 6 rows. We could dig a little deeper to identify more, but - we don't need to.

Since the CSV file has more accurate information, it's preferred. Let's juse use that.

```{r}
data <- data_csv
```

# Exploratory Data Analysis

The first thing I'd like to know is what the variables look like. 

## Non-Numeric Variables

For the non-numeric vars (like `Notes`) I'd like to know:

* How many unique vablues are there?
* What's the relative frequency?

```{r}
non_numeric_vars <- names(data)[!sapply(data, is.numeric)]
data %>% 
    select(one_of(non_numeric_vars)) %>%
    summarise_each(funs(unique_vars = length(unique(.))))
```

It looks like `EmployeeName` has the most unique values - in fact, relatively few people share names (~25%).

However, why are there so many values for `Benefits`? What's in that column?

```{r}
# No values in the first few rows - 
(head(data$Benefits))

# So how about unique values?
(head(unique(data$Benefits)))
```

Those are all numbers!

That makes some sense - by default `dplyr` scans the first 50 rows to infer type. If there were no numbers in the first 50 rows `dplyr` would cast to string.

When does the first non-empty value of benefits occur?

```{r}
which.min(data$Benefits == "")
```

That's quite a lot of empty rows!

Well, we can coerce to number.

```{r}
data <- data %>%
    mutate(Benefits = as.numeric(Benefits))

# Update the list of non-numeric vars
non_numeric_vars <- names(data)[!sapply(data, is.numeric)]
```

OK - so now that we have the refreshed list, let's look at the combinations of unique values in `Notes`, `Agency`, and `Status`.

```{r}
data %>%
    group_by(Notes, Agency, Status) %>%
    summarise(Frequency = n())
```

So - `Notes` contains no information; `Agency` is all the same; and `Status` is blank for most users. That looks unreliable... surely the only two options are *Part Time* (`PT`) or *Full Time* (`FT`)? 

For the moment let's remove `Notes` and `Agency` since they don't add any value. We'll figure out `Status` later.

```{r}
data <- data %>% 
    select(-Notes, -Agency)
```

### Inferring Gender

```{r}
## create a firstname column
data$fname<-sapply(strsplit(data$EmployeeName," "), "[[", 1)
```


### Job Title

I want to pay special attention to `JobTitle`. What information do we have?

Let's look at the most frequent.

```{r}
data %>%
    group_by(JobTitle) %>%
    summarise(Frequency = n()) %>%
    arrange(desc(Frequency)) %>%
    head
```

It looks like we could perhaps classify some of these, but that requires domain expertise (*Police Officer 3* is obviously Police, but do all Police job have Police in the title?)

How many job titles only occur once?

```{r}
data %>%
    group_by(JobTitle) %>%
    summarise(Frequency = n()) %>%
    mutate(OccursOnce = Frequency == 1) %>%
    group_by(OccursOnce) %>%
    summarise(Total = n())
```

OK, so most jobs occur more than once few times. What are some of the ones that occur only once?

```{r}
data %>% 
    filter(!duplicated(JobTitle)) %>%
    head()
```



They look pretty senior. How does the pay compare to the people who share a title?

```{r}
data %>%
    mutate(SharesTitle = duplicated(JobTitle)) %>%
    ggplot(aes(x = TotalPay)) + 
    geom_density(aes(fill = SharesTitle), alpha = 0.6)
```

Well, that's interesting.

First of all: it looks like people who don't share a title have somewhat higher pay - certainly in the right tail.

But why do so many people who *share* a title have such a high frequency of `TotalPay` at 0?

We obviously need to learn more about the numeric columns before embarking on this kind of analysis!

## Numeric Variables

Let's look at a summary of the numeric vars.

```{r}
data %>%
    select(-one_of(non_numeric_vars)) %>%
    summary
```

A few interesting things - 

* Some people have **negative** pay, . What on earth does that mean?
* `Year` looks like a factor indicating year.
* Does `Id` identify the row, or the employee?
    
Let's take that in random order.

Does `Id` identify the employee?

```{r}
table(duplicated(data$Id))
```

Nope, it identifies the row. That's frustrating - what if someone got a promotion in 2013? How can we track the change in their salary?

Who has negative pay?

```{r}
data %>%
    filter(TotalPay < 0) 
```

OK, just one person. Perhaps he mis-stated an invoice in the previous year.

Who has pay of 0?

```{r}
data %>% 
    filter(TotalPay == 0) 
```

OK, so we have 368 people. 

But - some of them are getting benefits. Perhaps they're retired, on disability, or something else?

Alright - so now, how many people have pay between $0.00 and $100.00?

```{r}
data %>%
    filter(TotalPay > 0, TotalPay < 100) 
```

We have 398 people. This probably includes people who were paid for, say, an hour of their time.

Now: does the `Status` column have anything to do with this? Perhaps Full Time employees have more complete information.

## Job Status & Total Pay

Let's take a quick look.

```{r}
data %>%
    ggplot(aes(x = TotalPay)) +
    geom_density(aes(fill = Status), alpha = 0.6)
```

Well, that makes a lot of sense. Part Time (`PT`) employees are compensated according to how much the city uses them - which is very little.

The story with blank `Status` is a little more complicated. Frankly, it looks like a lot of the blank entries should be full time; and some should be part time. We can see the graph split.

For now, let's proceed by filtering down to `FT` employees. Later, we might want to do something with the missing values in the `Status` column (for instance, predict the right value).

# Analyzing Full Time Employees

First let's create a dataset just for full time employees.

```{r}
ft <- data %>% filter(Status == "FT")
dim(ft)
```

OK. We have ~22,000 rows. 

How does that break down by each year?

```{r}
ft %>%
    group_by(Year) %>%
    summarise(Frequency = n())
```

They are **all** from 2014. Good to know. 

That introduces the question about the blank `Status` rows. Perhaps values are only filled out for 2014? Let's take a quick look before progressing.

```{r}
data %>%
    group_by(Year, Status) %>%
    summarise(Frequency = n()) %>%
    spread(Year, Frequency) 
```

Well, that answers that question: **only** 2014 has people categorized, and even 2014 has 4 missing values (!). Who are they?

```{r}
data %>%
    filter(Status == "", Year == 2014) 
```

That's... interesting. No names, and no pay? 

Anyway: back to 2014 Full Time employees.

## Compensation

To start with, let's look at compensation.

```{r}
ft %>%
    ggplot(aes(x = BasePay)) +
    geom_density(fill = "grey40") 
```

It looks like we have a few spikes there - perhaps diferent types of jobs, or different levels?

What about overtime?

```{r}
ft %>%
    ggplot(aes(x = OvertimePay)) +
    geom_density(fill = "grey40") 
```

That's a .... pretty weird graph. Someone made over $150,000 in *overtime*? Really?

OK, so how about benefits.

```{r}
ft %>%
    ggplot(aes(x = Benefits)) +
    geom_density(fill = "grey40") 
```

That looks a little better; again there are spikes. Interesting that the lowest benefit number is ~$25,000 annually, or $2,000/month. 

## Job Titles

OK: there are ~22,000 people doing full time work. How many unique job titles?

```{r}
# How many
(length(unique(ft$JobTitle)))

# Quick previous
head(unique(ft$JobTitle))
```

I'd really like to categorize these somehow. 

### Job Levels

Taking a quick look, it seems like some jobs end in a level (e.g. "2" or "I"). Let's take a look.

```{r}
# Add to dataset
ft %>%
    mutate(level = str_extract(JobTitle, "(?<=\\s).$")) ->
    ft

# Subset down
ft %>%
    filter(!is.na(level)) ->
    jobs_with_levels

# Peek
(head(jobs_with_levels))

jobs_with_levels %>%
    group_by(level) %>%
    summarise(People = n(), Titles = n_distinct(JobTitle))
    
```

So, it looks like we have ~150 of the job titles and ~3,500 people that have levels.

Well - how do the levels compare?

```{r}
ft %>%
    mutate(Level = ifelse(is.na(level), "No Level", level)) %>% 
    ggplot(aes(x = TotalPay)) +
    geom_density(aes(fill = Level), alpha = 0.6) +
    facet_wrap(~Level, scales="free_y")
```

Unsurprisingly, it looks like 1 veers lower. Interesting to see 2, 3, 4, and 5 - there look to be at least two underlying distributions in each.


### Title Level

What about the information in "Asst" and "Associate", etc?

Let's look at the first word.

```{r}
ft %>%
    mutate(first_word = str_extract(JobTitle, "^.+?(?=\\s)")) %>%
    group_by(first_word) %>%
    summarise(frequency = n()) %>%
    arrange(desc(frequency)) 
```

The `NA` are introduced because the job title is only one word.

Unfortunately, the most frequent ones don't pop out too well. 

### Cheating: Job Type

The SQL script [SF Salaries by JobType](https://www.kaggle.com/mannychapman/d/kaggle/sf-salaries/sf-salaries-by-jobtype) created by [mannychapman](https://www.kaggle.com/mannychapman) has some classification of job type. It'd be a shame to re-create work already done, so let's re-implement here.

Since I'm a lazy man, I'd going to re-use the script almost as-is. We have that nice SQLite database - why not use it to retrieve IDs?

```{r}
query = "select  Id, 
            case when jobtitle like '%fire%' then 'Fire'
            when jobtitle like '%police%' then 'Police'
            when jobtitle like '%sherif%' then 'Police'
            when jobtitle like '%probation%' then 'Police'
            when jobtitle like '%Sergeant%' then 'Police'
            
            
            when jobtitle like '%MTA%' then 'Transit'
            when jobtitle like '%transit%' then 'Transit'
            when jobtitle like '%anesth%' then 'Medical'
            when jobtitle like '%medical%' then 'Medical'
            when jobtitle like '%nurs%' then 'Medical'
            when jobtitle like '%health%' then 'Medical'
            when jobtitle like '%physician%' then 'Medical'
            when jobtitle like '%Orthopedic%' then 'Medical'
            
            when jobtitle like '%health%' then 'Medical'
            when jobtitle like '%pharm%' then 'Medical'
            when jobtitle like '%airport%' then 'Airport'
            when jobtitle like '%animal%' then 'Animal'
            when jobtitle like '%architect%' then 'Architectural'
            when jobtitle like '%court%' then 'Court'
            when jobtitle like '%legal%' then 'Court'
            
            when jobtitle like '%MAYOR%' then 'Mayor'
            when jobtitle like '%librar%' then 'Library'
            
            when jobtitle like '%parking%' then 'Parking'
            
            when jobtitle like '%Public Works%' then 'Public Works'
            when jobtitle like '%Attorney%' then 'Attorney'
            when jobtitle like '%MECHANIC%' then 'Automotive'
            when jobtitle like '%automotive%' then 'Automotive'
            when jobtitle like '%custodian%' then 'Custodian'
            when jobtitle like '%engineer%' then 'Engineering'
            when jobtitle like '%engr%' then 'Engineering'
            when jobtitle like '%account%' then 'Accounting'
            when jobtitle like '%Gardener%' then 'Gardening'
            when jobtitle like '%General Laborer%' then 'General Laborer'
            when jobtitle like '%food serv%' then 'Food Service'
            when jobtitle like '%clerk%' then 'Clerk'
            when jobtitle like '%porter%' then 'Porter'
            else null 
        end as JobType
        from Salaries"

db %>%
    tbl(sql(query)) %>%
    collect() ->
    job_classification

# Now we can join with ft
ft %>%
    left_join(job_classification) ->
    ft

# Let's see how many jobs were *not* matched.
ft %>% 
    filter(is.na(JobType)) %>%
    summarise(Rows = n(), JobTitles = n_distinct(JobTitle)) 
```

OK: so we missed around 633 of the `r length(unique(ft$JobTitle))` jobs.

Well, let's take a look at the results anyway.

```{r}
ft %>%
    ggplot(aes(x = TotalPayBenefits)) +
    geom_density(alpha = 0.6) +
    facet_wrap(~JobType, scales = "free_y", ncol = 4)
```

Well, that's a little interesting. Look at **Engineering** - is that 3 spikes? Also, steep ascent (looks like a minimum lower bound) as well as a pretty steep cliff at the high end. 

Perhaps the City needs to compete with tech companies at the low end (lower bar), and then doesn't compete at the higher end (so people leave?).

And look at **Attorney** - seems like a decent number start low, but most attorneys make the same amount of money.

**Medical** suggests there's more going on, with such a clear indication of two underlying distributions. 

Why don't we take a closer look at **Medical**?

#### JobType Medical

We might as well look at a Mixture Model for this.

First of all, let's take a closer look.

```{r}
ft %>%
    filter(JobType == "Medical") %>%
    ggplot(aes(x = TotalPayBenefits)) +
    geom_density() +
    scale_y_continuous(labels = dollar)
```

It looks like we have one peak around $100,000, and another just below $200,000.

Let's figure out roughly what they could be...

```{r}
ft %>%
    filter(JobType == "Medical") %>%
    mutate(breaks = TotalPayBenefits > 150000) %>%
    group_by(breaks) %>% 
    summarise(median = median(TotalPayBenefits),
              mean = mean(TotalPayBenefits),
              sd = sd(TotalPayBenefits))
```

Alright - let's fit a mixture model.

```{r}
medical <- filter(ft, JobType == "Medical")
normalmixEM(medical$TotalPayBenefits, lambda = .5, 
            mu = c(104383, 209014), sigma = c(18060,36480)) %>%
    plot(density = T)
```

That doesn't look unreasonable; though, of course, it's hard to identify which jobs would be in each category (or what the categories are) without more work. It could also be very wrong (and likely is: the second bump is probably senior staff, supervisors, management, etc).

### Job Types

I suspect we can do a better job. I'm not going to, since I'm about done with the time I set aside to do this work,  but let's consider:

* Assistants
* Supervisors

Take Assistants - 

```{r}
ft %>%
    mutate(Assistant = str_extract(JobTitle, "Asst|Assistant")) %>%
    group_by(Assistant) %>%
    summarise(mean(TotalPayBenefits))
```

It looks like people who are called **Asst** tend to make more money. It turns out - great surprise! - that when the City shortens the word, it's because it's a long title. Often when **Manager** or **Supervisor** is after it.

```{r}
ft %>%
    mutate(Assistant = str_extract(JobTitle, "Asst|Assistant")) %>%
    filter(Assistant == "Asst") %>%
    group_by(JobTitle) %>%
    summarise(Frequency = n(), pay = mean(TotalPayBenefits), sd = sd(TotalPayBenefits)) %>%
    arrange(desc(Frequency))
```

But, like anything, not always. Look at the **Asst Engr** ones! They drive the average up.

The compensation looks to be all over the place. Hmm. 

#### Word Compensation

What if we measured average compensation by word?

```{r}
words <- strsplit(tolower(ft$JobTitle), "\\W")
names(words) <- ft$Id

# Create tall matrix.
words_long <- data.frame(Id = names(unlist(words)), 
                         Word = unlist(words), 
                         stringsAsFactors = F)

# Strip off index position.
words_long <- words_long %>% 
    mutate(Id = gsub("\\d$","", Id),
           Id = as.integer(Id))

ft_comp <- ft %>% select(Id, JobTitle, TotalPayBenefits)

words_long %>% left_join(ft_comp) ->
    words_salary

head(words_salary)
```

OK, so just quickly - for every word that occurs at least twice, return the average.

```{r}
words_salary %>%
    group_by(Word) %>%
    summarise(Pay = mean(TotalPayBenefits, na.rm = T),
              sd = sd(TotalPayBenefits, na.rm = T),
              titles = n_distinct(JobTitle),
              people = n_distinct(Id)) %>%
    filter(people > 1, !is.na(Pay)) %>%
    mutate(Pay = dollar(Pay), sd = dollar(sd)) %>%
    arrange(desc(titles)) %>%
    head(20)
```

Alrighty then. Looks like there's some information there.

Now - just for fun - what happens if we make a model? Let's restruct to words that occur at least 20 times. We'll miss some (like Mayor), but might get some interesting results.

```{r}
words_to_model <- words_salary %>%
    group_by(Word) %>% 
    summarise(people = n_distinct(Id)) %>%
    filter(people >= 20) 

words_model <- subset(words_salary, Word %in% words_to_model$Word)

model <- lm(TotalPayBenefits ~ Word, words_model)
model_summary = summary(model)
```

First of all, what's our R squared as a base measure?

```{r}
model_summary$adj.r.squared
```

That's better than I expected!

So: what words are signifcant?

```{r}
coefficients = model_summary$coefficients
sig_coef = coefficients[coefficients[, 4] < 0.05, ] 
# Fix the names
rownames(sig_coef) <- gsub("^Word", "", rownames(sig_coef))
sig_coef <- as.data.frame(sig_coef, row.names = rownames(sig_coef))
sig_coef$Word = rownames(sig_coef)
head(sig_coef)
```

OK: the intercept is $166 - much higher than the actual average (~$130k)

What coefficients have the biggest negative effect?

```{r}
sig_coef %>%
    arrange(Estimate) %>% 
    head
```

What about the biggest positive effect?

```{r}
sig_coef %>%
    arrange(desc(Estimate)) %>% 
    head
```

Battalion? Really?

```{r}
ft %>%
    filter(grepl("battalion", tolower(JobTitle)))
```

Interesting - all inthe Fire Supress group I imagine that's Fire Department.

# Conclusion

Well, that's enough for now.

There are open challenges in:

* Classifying the JobsTitle.
* Classifying the EmployeeName (didn't even touch that).
* Classifying the FTE Status of people not in 2014.
* Looking at Overtime / Other (also didn't touch)